Towards atomistic resolution structure of phosphatidylcholine glycerol backbone and choline headgroup at different ambient conditions

### 2015-02-20 HGmodelMANUSCRIPT/HGmodel_draft3.tex
updates to HGmodel manuscript

\begin{abstract}
Phospholipids are essential building blocks of biological membranes.
Despite of vast amount of accurate experimental data, the atomistic resolution structures sampled by the glycerol backbone and choline headgroup
in phoshatidylcholine bilayers are not known. Atomistic resolution molecular dynamics simulations model 
would automatically resolve the structures giving an interpretation of experimental results, if the model
reproduced the experimental data. In the present work we simulate phosphatidylcholine (PC) lipid bilayers with 13 different
atomistic models, and we compare simulations with experiments (in fully hydrated conditions) in terms of C--H bond vector order
parameters for the glycerol backbone and choline headgroups. None of the current models is not accurately enough to resolve the structure with experimental accuracy.
However, closer inspection of three best performing models (CHARMM36, GAFFlipid and MacRog) suggest
that improvements in the sampled dihedral angle distributions would potentilly lead to the model which
would resolve the structure. Despite of the inaccuracy in the fully hydrated
structures, the response to dehydration, i.e. the tilting of the P--N vector more parallel to membrane normal, 
is qualitatively correct in all models. The CHARMM36 and MacRog models describe the interactions
between lipids and cholesterol better than the Berger/H\"oltje model.
This work has been, and continues to be, progressed and discussed through the blog: \url{nmrlipids.blogspot.fi}. 
Everyone is invited to join the discussion and make contributions through the blog. 
The manuscript will be eventually submitted to an appropriate scientific journal. 
Everyone who has contributed to the work through the blog will be offered 
coauthorship. For more details see: \url{nmrlipids.blogspot.fi}.

@@

This work has been done as an open collaboration by using \url{nmrlipids.blogspot.fi} as a communication platform.
The approach is inspired by the Polymath project~\cite{gowers09}, however there are some essential differences. 
The manuscript pointing out problems in the glycerol backbone and
headgroup structure in the most used molecular dynamics simulation model for lipid bilayers (Berger based models) was used as a starting point~\cite{ollila13}.
After publishing the initial manuscript, the blog was opened for further contributions and discussion from anyone interested. 
All the contributions were done publicly through the blog and the blog contributors were offered coauthorship. The final author list
is based on the self-assesment of the authors' scientific contribution to the project.

Large portion of the used simulation files and scripts are shared through the GitHub organization \url{https://github.com/NMRlipids}
and the Zenodo community \url{https://zenodo.org/collection/user-nmrlipids}.

@@

CHARMM36~\cite{klauda10}       & DPPC   & 72  &  2189 & 323 & 30 & 25  & ?\todoi{Samuli, put to Zenodo}  & SI \\

@@

Chiu \cite{chiu09} \todoi{Samuli, put to Zenodo}     & POPC  & 128 & 3552  & 298 & 56 & 50  & ? & SI \\

@@

In general, we conclude that the atomistic resolution classical molecular dynamics simulations 
is extremely convenient tool to give structural interpretation for the high resolution NMR data~\cite{ferreira14}. 
However, in the case of phoshatidylcholine glycerol backbone and choline headgroup further model development is still required.

This work has been, and continues to be, progressed and discussed through a blog at \url{nmrlipids.blogspot.fi}. 
Everyone is invited to join the discussion and make contributions through the blog. 
The manuscript will be eventually submitted to an appropriate scientific journal. 
Everyone who has contributed to the work through the blog will be offered 
coauthorship. For more details, see \url{nmrlipids.blogspot.fi}.

@@

\subsection{General issues to think of about the manuscript}
\todo{Should we emphasize more the excellent collection of simulation trajectories collected to Zenodo?} \\
\todo{Maybe we should mention that CHARMM is significanlty more computationally expensive than other all-atom models.
This seems evident for a person who has ran CHARMM and other all-atom models with the same computer. However, there
may be some people who have never used any other model than CHARMM. In this case it may be difficult to realize.
Would anyone have a good bencmark about this? I have assumed that the reason for the slowness are the extra Lennart-Jones
interactionsi in water, however, I have never looked into it. Would someone have some hard facts on this?}

>>>>

\begin{abstract}
Phospholipids are essential building blocks of biological membranes.
Despite of vast amount of accurate experimental data, the atomistic resolution structures sampled by the glycerol backbone and choline headgroup
in phoshatidylcholine bilayers are not known. Atomistic resolution molecular dynamics simulations model 
would automatically resolve the structures giving an interpretation of experimental results, if the model
reproduced the experimental data. In the present work we simulate phosphatidylcholine (PC) lipid bilayers with 13 different
atomistic models, and we compare simulations with experiments (in fully hydrated conditions) in terms of C--H bond vector order
parameters for the glycerol backbone and choline headgroups. None of the current models is not accurately enough to resolve the structure with experimental accuracy.
However, closer inspection of three best performing models (CHARMM36, GAFFlipid and MacRog) suggest
that improvements in the sampled dihedral angle distributions would potentilly lead to the model which
would resolve the structure. Despite of the inaccuracy in the fully hydrated
structures, the response to dehydration, i.e. the tilting of the P--N vector more parallel to membrane normal, 
is qualitatively correct in all models. The CHARMM36 and MacRog models describe the interactions
between lipids and cholesterol better than the Berger/H\"oltje model.
{\bf This work has been done as an open collaboration by using the \url{nmrlipids.blogspot.fi} as an
communication platform. All the scientific contributions have been done through the blog and
are publicly available. In addition, almost all simulation trajectories and files are made available
in the Zenodo community \url{https://zenodo.org/collection/user-nmrlipids} which has became the most extensive
publicly available collection of molecular dynamics simulation trajectories of lipid bilayers.
}
%This work has been, and continues to be, progressed and discussed through the blog: \url{nmrlipids.blogspot.fi}. 
%Everyone is invited to join the discussion and make contributions through the blog. 
%The manuscript will be eventually submitted to an appropriate scientific journal. 
%Everyone who has contributed to the work through the blog will be offered 
%coauthorship. For more details see: \url{nmrlipids.blogspot.fi}.

@@

This work has been done as an open collaboration by using \url{nmrlipids.blogspot.fi} as a communication platform.
The approach is inspired by the Polymath project~\cite{gowers09}, however there are some essential differences. 
{\bf The present approach was started by publishing a manuscript~\cite{ollila13} discussing the glycerol backbone and choline structures 
in a Berger based model (the most used molecular dynamics simulation model for lipid bilayers).
After publishing the initial manuscript, the open invitation for further contributions and discussion through the blog was presented.
All the scientific contributions are done publicly through the blog and all contributors were offered coauthorship. 
The acception of authorship is based on the self-assesment of the authors' scientific contribution to the project.
The contributions from each author are summarized in ?? (the location depends on the journal probably).

Almost all simulation data, including running parameters for reprocution and trajectories for further analysis, are collected
to the nmrlipids community in Zenodo file repository \url{https://zenodo.org/collection/user-nmrlipids}.
Thus, besides the main topic of this manuscript we present the most extensive available collection of simulations trajectories
for lipid bilayers, which opens up numerous possiblities for different analyses with much less effort than previously required.
Furher information, e.g. scripts, figures and manuscript text files, are available in the GitHub organization \url{https://github.com/NMRlipids}.
}

@@

CHARMM36~\cite{klauda10}       & DPPC   & 72  &  2189 & 323 & 30 & 25  & \cite{charmmFILESdppc}$^*$\todoi{DONE}  & SI \\

@@

Chiu \cite{chiu09} \todoi{DONE}     & POPC  & 128 & 3552  & 298 & 56 & 50  & \cite{chiuFILES}$^*$ & SI \\

@@

- {\bf Besides the main conclusions, we have created the most extensive
publicly available collection of molecular dynamics simulation trajectories of lipid bilayers
into the Zenodo community \url{https://zenodo.org/collection/user-nmrlipids}. 
This collection opens up numerous possiblities for different analyses with
much less effort than previously required.

In general, we conclude that to fully utilize the potential of atomistic resolution classical molecular dynamics simulations
in the structural interpretation of the high resolution NMR data~\cite{ferreira14} for lipid bialyers one has to  
improve the phoshatidylcholine glycerol backbone and choline headgroup parameters.

This work has been done as an open collaboration by using the \url{nmrlipids.blogspot.fi} as an communication
platform. All the scientific contributions have been done through the blog and are publicly
available. }

%This work has been, and continues to be, progressed and discussed through a blog at \url{nmrlipids.blogspot.fi}. 
%Everyone is invited to join the discussion and make contributions through the blog. 
%The manuscript will be eventually submitted to an appropriate scientific journal. 
%Everyone who has contributed to the work through the blog will be offered 
%coauthorship. For more details, see \url{nmrlipids.blogspot.fi}.

@@

\subsection{General issues to think of about the manuscript}
\todo{DONE}\\
\todo{Maybe we should mention that CHARMM is significanlty more computationally expensive than other all-atom models.
This seems evident for a person who has ran CHARMM and other all-atom models with the same computer. However, there
may be some people who have never used any other model than CHARMM. In this case it may be difficult to realize.
Would anyone have a good bencmark about this? I have assumed that the reason for the slowness are the extra Lennart-Jones
interactionsi in water, however, I have never looked into it. Would someone have some hard facts on this?}
